\section{Information flows in the DGA}
\label{sec:dga_flows}

In this Section, a detailed description of the DGA and its core legislative objectives is provided, including the introduction of use cases where Semantic Web technologies and decentralised data environments can be leveraged to aid data intermediaries and data altruism organisations to implement their services while fulfilling their renewed legal duties, in particular, related to the reporting of their activities and how they use personal and non-personal data from data subjects and data holders, respectively.

In November 2020, the \cite{european_commission_communication_2020} introduced a series of regulatory proposals aimed at legislating the European strategy for data.
Among these proposals was the Data Governance Act \citeyearpar{noauthor_regulation_2022}, a regulation presented to enhance data accessibility and foster confidence in data intermediation services throughout the EU.
Following approval by both the European Parliament and the European Council, this legislation entered into force on 23 June 2022 and, following a 15 month grace period, has been applicable since September 2023.
Similarly to other data-related legislation within the EU, the DGA relays new rights and obligations to entities holding both personal and non-personal data.
It also regulates the operations of data users and two categories of data-related services related to data intermediation and altruism.
The principal objectives of this legislation include:
\begin{enumerate}
    \item Facilitating the reuse of protected public-sector data while maintaining its privacy and confidentiality, particularly in cases where such data is subject to the rights of others, including trade secrets, personal data protection, and data safeguarded by intellectual property rights.
    \item Regulating and maintaining a registry of data intermediation service providers, which facilitate data sharing among enterprises and support individuals to have a ``personal data-sharing intermediary'', designed to aid them in exercising their rights, e.g., under the GDPR.
    \item Allowing businesses and data subjects to voluntarily contribute data for altruistic purposes, such as medical research.
    \item Establishing a novel supervisory authority, the European Data Innovation Board, tasked with supervising the operations of data intermediation service providers and data altruism organisations.
\end{enumerate}

% TODO: definition of new entities: public sector bodies

\subsection{Reuse of protected data held by public sector bodies}
\label{sec:reuse}

DGA's Chapter II legislates the reuse of data stored by public sector bodies.
To keep such data, these entities need to have in place safeguards to protect their commercial and statistical confidentiality, intellectual property rights of third parties, and personal data-related rights.
Moreover, they have to publish the dataset reuse conditions, and the respective data request procedure, in a transparent and publicly accessible manner.
The reuse must be contracted, with a maximum duration of one year, including the categories of data being used as well as the purpose for the reuse. 
Public sector bodies also have the right to obtain guidance and technical support from a competent body, which must be appointed by each EU member state.
These appointed entities are responsible for providing guidance on data formats and storage, assisting in the implementation of privacy-preserving methods to protect personal data integrity, and supporting activities to obtain consent from data subjects and permission from data holders.

\subsection{Data intermediaries}
\label{sec:intermediation}

DGA is also the first of its kind to regulate the activity of data intermediation service providers.
Such entities \textit{``aim[s] to establish commercial relationships for the purposes of data sharing between an undetermined number of data subjects and data holders on the one hand and data users on the other, through technical, legal or other means''} \citeyearpar{noauthor_regulation_2022}.
Article 12 includes a list of conditions for the provision of this service, e.g., providers should have tools to convert data into specific formats, use standards to promote interoperability across sectors, gather data subjects' consent and data holders' permission terms, as well as update or withdraw these terms, and maintain records of their activity.
Before providing this service, intermediation providers must be registered in the public register maintained by the EC, and their activity must be supervised at the national level by a nominated competent authority.

\subsection{Data altruism}
\label{sec:altruism}

Data altruism as a term is also introduced by the DGA.
This term relates to the sharing of personal and non-personal data, based on data subjects' consent and data holders' permission, for the `common good'.
This entails purposes such as improving healthcare, combating climate change, or performing scientific research.
Each EU member state can establish its altruism policy and has to appoint a competent authority to oversee the activity of altruistic organisations (it can be the same as the one that supervises intermediation providers).
Said authority also has to keep up to date, public registries of altruistic organisations, including details on their identity, contact information, legal status, and main goals.
Data altruism organisations themselves have to maintain records of their activities, in particular, to produce annual reports to share with the relevant competent authority.
To facilitate this activity, a European data altruism consent form will be facilitated \beatriz{by the EC} to \textit{``allow the collection of consent or permission across Member States in a uniform format''}.

% Figure of information flows and documents (from the paper)
% Comparison with GDPR flows
% Identification of important use cases that can be tackled and easily extended with the work already developed for GDPR