% called by main.tex
%
\section*{Acknowledgement}
\label{sec::acknowledgement}
\addcontentsline{toc}{section}{Acknowledgement}

This thesis has been accomplished with the support, assistance and encouragement of many people to whom I would like to express my deepest gratitude.

First of all, I would like to express my gratitude to the Ontology Engineering Group for supporting the work developed in this thesis, in particular to my supervisor Professor Víctor Rodríguez Doncel.
Thank you for providing me with the incredible opportunity to explore and develop my own ideas in distinct fields of knowledge and for supporting me throughout this remarkable experience.
I would also like to express my gratitude to Professor Elena Montiel Ponsoda for the insightful discussions and the opportunity to work in INESData.
Furthermore, I would like to thank my co-supervisor, Professor Dave Lewis, for his enthusiasm, interest, and comprehensive support.
Last, but certainly not least, my sincere gratitude goes to Professor Harshvardhan Pandit for being an incredible mentor.
Thank you Harsh, for your assistance, discussions, collaborations, technical and legal expertise, as well as just chatting over many offline and online meetings.

Secondly, I would also like to acknowledge the funding received from the European Union's Horizon 2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement No 813497 (PROTECT), European Union’s Horizon 2020 research and innovation programme under grant agreement No. 101036418 (AURORA), and the NextGenerationEU-funded INESData project.
I would like to thank all the fellows and professors involved in these projects.
In particular, I am especially grateful to my fellow PROTECT Work Package 1 colleagues, Andrés Chomczyk Penedo, Blessing Mutiro, and Haleh Asgarinia, for the insightful discussions that contributed to the interdisciplinarity tone of this thesis.
It was, and still is, a pleasure to work with you.

I would also like to thank the people that made my research stays memorable and kickstarted collaborations that I hope will last for many years.
To Dr. Kieran Fraser, my research stay at Empushy was a great opportunity to apply my legal knowledge in a startup environment.
To Pat McBennett, our continuous discussions and interactions were immensely helpful in applying this thesis outcomes to relevant industrial settings.
To Dave and Harsh, my research stay at Trinity College Dublin was pivotal in the development of PLASMA and DGAterms.
To Professor Ruben Verborgh, while short, my research stay at UGent was incredibly insightful.
The work being developed in Flanders, centred on Solid, and in particular by the KNoWS group was and is an inspiration that drove the final touches of this thesis.
I would also like to extend a final thanks to Ruben for being incredibly kind and providing actionable feedback that immensely contributed to the final text that can be read in this thesis.

To the DPVCG and ODRL CG people, your incentive and feedback were immensely helpful for the validation of this thesis.

À Carina, obrigada pelo suporte, companheirismo e pelas muitas noites passadas a escrever papers e a fazer-me figuras.
Nada disto teria sido possível sem ti, obrigada por nunca me teres deixado desistir.
\textit{My personal assistant, forever.}
Por fim, não posso deixar de agradecer à minha família e amigos: 
mãe, pai, Nuno, avós, padrinho e madrinha, Gui e Afonso, e todos os outros, muito obrigada!
Obrigada a todos pelo apoio, pelo interesse e pela preocupação.